\vspace{1.0cm} \LARGE \bf fluB

### 2013-08-28 fluB.tex
Adding multiple files, minor adjustments to text.
Added poster version of the manuscript, acknowledgement tables (.tsv and
.xls formats) and added a few references to the manuscript in
preparation for ploughing through the manuscript once I'm back.

Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains. 
The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000}), thus limiting the virus's available antigenic reservoir. 
It has been suggested that influenza B viruses use a combination of reassortment and point mutations to generate antigenic diversity (comment - citation).

@@

\subsection*{Building trees}
\tbc{I generally prefer to describe the model (HKY, etc...) then say that posterior samples were drawn using MCMC and describe sampling regime.}

Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting (comment - reference robust counting) and GMRF Bayesian skyride (comment - reference skyride) as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) \tbc{CITE}.
We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)). Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e. Y to V or V to Y) between a node and its descendant in any segment. 
In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.

@@

\subsection*{Measures of diversity}
We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on diversity existing at each time period.
In addition, we calculated mean pairwise diversity and mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.

@@

(comment - END of the newer version)

\subsection*{Reassortment in influenza B is common and extensive}
Extensive reassortment has long been established as a major pathway by which influenza B viruses generate diversity.
Even focusing on a specific subset of reassortments (those that occur between two lineages and those where the previous lineage was of the opposite lineage) reveals 116 events and if we naively assume that all events are replicated across all trees it still means there have been over 14 inter-lineage reassortment events.
We also see that reassortments occur across considerable phylogenetic distances and some segment lineages have been so successful as part of reassortant viruses that they out-competed their sister lineages (e.g. Yamagata lineages of PA, NP, MP and NA segments). (comment - could include diversity over time plots to show maximum phylogenetic distance attained prior to reassortment for each segment)

\subsection*{Selection maintains the diversity in PB1-PB2-HA complexes of Victoria and Yamagata lineages}
Rather surprisingly, 3 segments have maintained a ratio of Vic and Yam lineages very close to 0.5 (PB1, PB2 and HA) (see Figure \ref{lineageRatios}) and have preserved the diversity that has accumulated since the split of the two lineages (see Figure \ref{tmrcaOT}).
Our results suggest that this is due to the 3 segments co-reassorting (comment - refer to posterior distributions of reassortant dates plot).
We find evidence that this co-reassortment pattern between PB1, PB2 and HA is not due to skewed reassortment (Marhsall \textit{et al.} \cite{marshall2013} have shown that reassortment efficiency is high in influenza A), but rather due to viruses bearing mixed-lineage PB1-PB2-HA complex having lower fitness than viruses with PB1-PB2-HA complexes derived entirely from Yam or Vic lineages (comment - refer to stratified REPU plot).
Not only are events that split up pure-lineage PB1-PB2-HA complexes rare, but they also have poor persistence and trees have spent less than 10\% of the time in the mixed-lineage PB1-PB2-HA complex state (Figure \ref{stateTime}), suggesting that they are selected against (comment - make plot of that).

(comment - PB1, PB2 and HA are missing extra years in the mixed-lineage bar due to invisible reassortments e.g. VVY is invisible in HA tree, YVV in PB1 and YVY in PB2; solution: remove PB1, PB2 and HA bars or make it clear in the text)

Thus our findings suggest that the link maintaining PB1, PB2 and HA lineages close to equilibrium is not due to biased reassortment, but a result of selection acting upon reassortants containing mixed-lineage PB1-PB2-HA segments.
We find that by 1995 sufficient co-dependence had developed between PB1, PB2 and HA segments of Victoria and Yamagata lineages to prevent reassortant progeny containing mixed-lineage PB1-PB2-HA complexes from infecting sufficient numbers of individuals to cross the detection threshold.

\subsection*{Segments vary in their abilities to reassort and success as part of reassortant viruses}
A rather striking pattern in influenza B evolution is the fact that segments reassorted at the same time can have dramatically different fates.
For example, PA, NP, MP and NS segments of Yamagata lineage were reassorted into Victoria lineage PB1-PB2-HA background in 1994, but of those 4 segments only the NP lineage persisted for a considerable amount of time, being replaced by another NP lineage in 2006, whereas PA was replaced in 2000 and MP together with NS were replaced by different lineages in 2004.
This suggests that most non-PB1, PB2 or HA segments have the capacity to evolve somewhat independently of the rest of the genome.

\subsection*{Interpreting our results}
(comment - this bit is exceedingly unsubstantiated and ramble-y)
Recently the antigenic drift of influenza B viruses has been characterised (comment - cite Bedford et al 2013) and Victoria lineage HA was found to drift nearly twice as fast as Yamagata lineage HA.
A possible explanation for this would be differences in polymerase fidelity: if Victoria lineage polymerase introduced higher mutation rates, this would explain the higher antigenic drift rate seen in Vic lineage HA, but also would suggest that Vic lineage PB1-PB2-HA complex "borrows" segments from Yamagata lineage PB1-PB2-HA complex-bearing viruses and "breaks" them through going over something akin to each segment's error threshold, necessitating another round of reassortment.
However, we find no evidence of differences in nucleotide, synonymous and non-synonymous rates of evolution between branches that were inferred to have been in the same virion as PB1, PB2 or HA segments of Vic or Yam lineages.
Thus we have no evidence to suggest fundamental differences between Vic and Yam lineage PB1-PB2-HA complexes.

Balancing selection would be expected, and does in the malaria parasite, preserve antigenic diversity, but it is unclear if maintenance of PB1-PB2-HA lineages is driven by balancing selection acting to preserve antigenic diversity of HA or the diversity of PB1 and PB2 segments.
Recently a similar link has been described in influenza A by Cobbin \textit{et al.} \cite{cobbin2013}, where the PB1 segment is reassorted more frequently than expected during vaccine strain creation.
If influenza B virus PB1 has a similar gene-regulatory activity, then it would be reasonable to assume that HA and PB1 are associated through the necessity for HA to have appropriate levels of expression and PB2 hitchhikes together with PB1 due to some functional requirement as part of the polymerase complex.
This does not explain why the Vic lineage of the PA segment was lost so readily, however, despite it being a part of the polymerase complex.

\subsection*{Future evolution of influenza B viruses}
From our current findings it is difficult to extropolate what reassortants will come to dominate the influenza B population in the far future.
It seems likely that PB1, PB2 and HA segments of both Victoria and Yamagata lineages will either continue to co-exist in their co-adapted state or one lineage will go extinct.
Until the 2004 and 2006 reassortment events it seemed like the NP segment would follow the PB1-PB2-HA complex in becoming the next segment to be recruited as part of a co-adapted segment complex.
This method of gradually recruiting more segments to a co-adapted segment complex could have lead to the eventual speciation of influenza B into originally Vic and Yam lineage HA bearing species.
As it stands now, however, the most diversity between Victoria and Yamgata PB1-PB2-HA lineage-bearing viruses has been built up in PA and NA segments (which shared a common ancestor around 2000) and it remains to be seen whether this diversity will too be lost following a reassortment event.

Given the evolutionary history of the NP segment, it would be reasonable to focus sequencing efforts on complete influenza B genome sequencing.
Our results indicate that successful reassortants are likely to come from the combination of Victoria lineage PB1-PB2-HA complex and PA, NP, NA and MP segments of Yamagata lineage (or NS segment of Vic lineage), though the number of Yamagata lineage segments can vary.
It is also obvious that increased surveillance efforts will inevitably uncover a lot of reassortant viruses (possibly even viruses with mixed-lineage PB1-PB2-HA complexes) and the threshold at which these reassortants should be considered as being likely to take over the influenza B population would require further work.

\subsection*{Testing the co-dependence between PB1, PB2 and HA segments}
Our results suggest a simple test for the hypothesis that PB1, PB2 and HA segments of Victoria and Yamagata lineages have become co-dependent - if PB1, PB2 or HA segments from different lineages were combined artificially in a single virion, these should have lower fitness than virions containing wholly Victoria or wholly Yamagata lineage PB1-PB2-HA complex.
We also expect that the fitness of artificially derived reassortants would decline with increasing date of strain isolation i.e. mixed-lineage PB1-PB2-HA containing reassortants with segments from older isolates having higher fitness than reassortants possessing segments of more recent isolates.

\subsection*{Implications for influenza B surveillance and prevention}
The findings presented here have several implications.
On one hand, if the co-dependence between PB1, PB2 and HA segments is strong enough, influenza B viruses containing mixed-lineage PB1-PB2-HA complex detected by surveillance are unlikely to contribute to long-term evolution of influenza B viruses.
On the other hand, the dependence of the HA segment on PB1 and PB2 segments (or \textit{vice versa}) implies that there is a possibility that more subtle disease control measures could be developed and used against influenza B viruses, or, if similar co-adapted segment complexes were discovered in influenza A, seasonal influenza in general.
In addition, the extent of inter-lineage reassortment argues for more intensive surveillance of influenza B viruses, as sequences of a handful of segments will not necessarily tell the whole story, especially since influenza B viruses have been known to evolve counter-intuitively (\textit{e.g.} losing Victoria lineages of PA and NA segments despite associations with the rest of the polymerase complex or antagonising HA activity, respectively).
Finally, this study also highlights the gaps in our knowledge about the molecular biology of influenza B viruses.

>>>>

Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains. 
The only known persistent source of influenza B viruses are humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), thus limiting the virus's available antigenic reservoir. 
It has been suggested that influenza B viruses use a combination of reassortment and point mutations to generate antigenic diversity \cite{nerome1998,mccullers1999}.

@@

\subsection*{Building trees}
Temporal phylogenies were recovered for each segment using the HKY model of nucleotide substitution, 3 codon partitions, robust counting (comment - reference robust counting) and GMRF Bayesian skyride (comment - reference skyride) as the demographic model over the course of an MCMC chain (3 parallel runs of 200 million states, sampling every 20000 steps).
In addition we used `relaxed' tips for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) \tbc{CITE}.
We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS (comment - cite Philippe)). Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e. Y to V or V to Y) between a node and its descendant in any segment. 
In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.

@@

\subsection*{Measures of diversity}
We inferred the diversity of each segment over time by finding the oldest most recent common ancestor of all branches existing at yearly intervals, which places an upper boundary on diversity existing at each time period.
In addition, we calculated mean pairwise TMRCA between branches inferred to be in virions possessing Vic and Yam lineage PB1, PB2 and HA.

@@
